The Unicode Cookbook \linebreak for Linguists

### 2016-06-30 book/unicode.tex
adding R examples

\include{chapters/preface}
\include{chapters/introduction}
\include{chapters/pitfalls}
\include{chapters/ipa_background}
\include{chapters/ipa_meets_unicode}

\include{chapters/orthography_profiles}

\section{Available implementations}
\label{implementations}

To illustrate the practical application of orthography profiles, we have
implemented two different versions of the above specifications, one in Python
and one in R. These two implementations have rather different implementation
histories, and we are not 100\% sure that they will in all situations give the
same results. Also the perfomance with larger datasets differs, and the code is
not always as clean as we would like it to be. In sum, the two implementations
should be considered as `proof of concept' and not as the final word on the
practical application of the specifications above. In our own experience, the
current implementations are sufficiently fast and stable to be useful for
academic practice (e.g. checking data consistency, or analyzing and
transliterating small to medium sized data sets), but they should probably
better not be used for full-scale industry applications.

@@

\noindent Basically, the idea is to use \texttt{write.profile} to produce a
basic orthography profile from some data and then \texttt{tokenize} to apply the
(possibly edited) profile on some data, as exemplified in the next section. This
can of course be performed inside R, but additionally there are two more
interfaces to the R code supplied in the \texttt{qlcData} package: (i) bash
executables and (ii) `shiny' webapps.

@@

\noindent For example, a good way to use the executables in a terminal is to
make softlinks (using \texttt{ln}) from the executables to a directory in your
PATH, e.g. to \texttt{/usr/local/bin/}. The two executables are named
\texttt{tokenize} and \texttt{writeprofile}

\begin{knitrout}\scriptsize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\noindent
\ttfamily
\hlstd{}\hlslc{\#\ make\ two\ softlinks\ to\ the\ R\ executables\ in\ bash}\hspace*{\fill}\\
\hlstd{}\hlkwc{ln\ }\hlstd{}\hlopt{{-}}\hlstd{is\ }\hlstr{`Rscript\ {-}e\ 'cat(file.path(find.package("qlcData"),}\hspace*{\fill}\\
\hlstr{\ "exec",\ "tokenize"))'`}\hlstd{\ }\hlopt{/}\hlstd{usr}\hlopt{/}\hlstd{}\hlkwb{local}\hlstd{}\hlopt{/}\hlstd{bin}\hspace*{\fill}\\
\hlkwc{ln\ }\hlstd{}\hlopt{{-}}\hlstd{is\ }\hlstr{`Rscript\ {-}e\ 'cat(file.path(find.package("qlcData"),}\hspace*{\fill}\\
\hlstr{\ "exec",\ "writeprofile"))'`}\hlstd{\ }\hlopt{/}\hlstd{usr}\hlopt{/}\hlstd{}\hlkwb{local}\hlstd{}\hlopt{/}\hlstd{bin}\hspace*{\fill}
\mbox{}
\normalfont
\end{kframe}
\end{knitrout}

\noindent After this addition it should be possible to access the \texttt{tokenize}
function from the shell. Try \texttt{tokenize --help} to test the functionality.

\begin{knitrout}\tiny
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\noindent
\ttfamily
\hlstd{tokenize\ }\hlopt{{-}{-}}\hlstd{}\hlkwb{help}\hlstd{}\hspace*{\fill}
\mbox{}
\normalfont

@@

\section{Abstract examples}

\subsection*{Error reporting}

@@

\noindent The working of error-message reporting can also nicely be illustrated with this
example. Suppose we made an orthography profile with just the Latin <A> and <B> 
as possible graphemes, then this profile would not be sufficient to tokenize the 
strings. There are graphemes in the data that are not in the profile, so the 
tokenization produces an error, which can be used to fix the encoding. In the 
example below, we can see that the Cyrillic encoding is found in the second 
string of the \texttt{test} input.

@@

\subsection*{Different ways to write a profile}

The preparation of an orthography profile from some data might sound like a 
trivial problem, but actually there are various different ways in which strings 
can be separated into graphemes by \texttt{write.profile}. Consider the 
following example string. The default settings of \texttt{write.profile} 
separates the string into Unicode graphemes according to grapheme clusters 
(`user-perceived characters', see Section~\ref{the-unicode-approach}).

@@

% latex table generated in R 3.3.0 by xtable 1.8-2 package
% Wed Jun 22 12:03:05 2016
\begin{table}[H]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Gr. & Freq. & Codepoint & Unicode Name \\ 
  \midrule
Ú & 1 & U+00DA & LATIN CAPITAL LETTER U WITH ACUTE \\ 
  Ú & 1 & U+0055, U+0301 & LATIN CAPITAL LETTER U, COMBINING ACUTE ACCENT \\ 
  Ù & 1 & U+00D9 & LATIN CAPITAL LETTER U WITH GRAVE \\ 
  Ù & 1 & U+0055, U+0300 & LATIN CAPITAL LETTER U, COMBINING GRAVE ACCENT \\ 
  Û & 1 & U+00DB & LATIN CAPITAL LETTER U WITH CIRCUMFLEX \\ 
  Û & 1 & U+0055, U+0302 & LATIN CAPITAL LETTER U, COMBINING CIRCUMFLEX ACCENT \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Profile 1 (default settings, splitting grapheme clusters)} 
\label{tab:profile1}
\end{table}

\noindent Some of these graphemes are single codepoints, others are combinations
of two codepoints. By specifying the separator as empty, it is possible to split
the string into codepoints.

@@

% latex table generated in R 3.3.0 by xtable 1.8-2 package
% Wed Jun 22 12:03:05 2016
\begin{table}[H]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Grapheme & Frequency & Codepoint & Unicode Name \\ 
  \midrule
́ & 1 & U+0301 & COMBINING ACUTE ACCENT \\ 
  ̀ & 1 & U+0300 & COMBINING GRAVE ACCENT \\ 
  ̂ & 1 & U+0302 & COMBINING CIRCUMFLEX ACCENT \\ 
  U & 3 & U+0055 & LATIN CAPITAL LETTER U \\ 
  Ú & 1 & U+00DA & LATIN CAPITAL LETTER U WITH ACUTE \\ 
  Ù & 1 & U+00D9 & LATIN CAPITAL LETTER U WITH GRAVE \\ 
  Û & 1 & U+00DB & LATIN CAPITAL LETTER U WITH CIRCUMFLEX \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Profile 2 (splitting by codepoints)} 
\label{tab:profile2}
\end{table}

@@

% latex table generated in R 3.3.0 by xtable 1.8-2 package
% Wed Jun 22 12:03:05 2016
\begin{table}[H]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Grapheme & Frequency & Codepoint & Unicode Name \\ 
  \midrule
Ú & 2 & U+00DA & LATIN CAPITAL LETTER U WITH ACUTE \\ 
  Ù & 2 & U+00D9 & LATIN CAPITAL LETTER U WITH GRAVE \\ 
  Û & 2 & U+00DB & LATIN CAPITAL LETTER U WITH CIRCUMFLEX \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Profile 3 (splitting by NFC codepoints)} 
\label{tab:profile3}
\end{table}

% latex table generated in R 3.3.0 by xtable 1.8-2 package
% Wed Jun 22 12:03:05 2016
\begin{table}[H]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Grapheme & Frequency & Codepoint & Unicode Name \\ 
  \midrule
́ & 2 & U+0301 & COMBINING ACUTE ACCENT \\ 
  ̀ & 2 & U+0300 & COMBINING GRAVE ACCENT \\ 
  ̂ & 2 & U+0302 & COMBINING CIRCUMFLEX ACCENT \\ 
  U & 6 & U+0055 & LATIN CAPITAL LETTER U \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Profile 4 (splitting by NFD codepoints)} 
\label{tab:profile4}
\end{table}

@@

\begin{knitrout}\footnotesize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# note that NFC and NFD normalization are identical for unicode grapheme}
\hlcom{# definitions}
\hlstd{profile_5} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(example,} \hlkwc{normalize} \hlstd{=} \hlstr{"NFD"}\hlstd{)}
\hlstd{profile_6} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(example,} \hlkwc{normalize} \hlstd{=} \hlstr{"NFC"}\hlstd{)}
\end{alltt}
\end{kframe}
\end{knitrout}

% latex table generated in R 3.3.0 by xtable 1.8-2 package
% Wed Jun 22 12:03:05 2016
\begin{table}[H]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Gr. & Freq. & Codepoint & Unicode Name \\ 
  \midrule
Ú & 2 & U+0055, U+0301 & LATIN CAPITAL LETTER U, COMBINING ACUTE ACCENT \\ 
  Ù & 2 & U+0055, U+0300 & LATIN CAPITAL LETTER U, COMBINING GRAVE ACCENT \\ 
  Û & 2 & U+0055, U+0302 & LATIN CAPITAL LETTER U, COMBINING CIRCUMFLEX ACCENT \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Profile 5 (splitting by graphemes after NFD)} 
\label{tab:profile5}
\end{table}

% latex table generated in R 3.3.0 by xtable 1.8-2 package
% Wed Jun 22 12:03:05 2016
\begin{table}[H]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Gr. & Freq. & Codepoint & Unicode Name \\ 
  \midrule
Ú & 2 & U+00DA & LATIN CAPITAL LETTER U WITH ACUTE \\ 
  Ù & 2 & U+00D9 & LATIN CAPITAL LETTER U WITH GRAVE \\ 
  Û & 2 & U+00DB & LATIN CAPITAL LETTER U WITH CIRCUMFLEX \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Profile 6 (splitting by graphemes after NFC)} 
\label{tab:profile6}
\end{table}

@@

\begin{knitrout}\tiny
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\noindent
\ttfamily
\hlstd{writeprofile\ }\hlopt{{-}{-}}\hlstd{}\hlkwb{help}\hlstd{}\hspace*{\fill}
\mbox{}
\normalfont

@@

>>>>

\include{chapters/preface}
\include{chapters/introduction}
\include{chapters/pitfalls}
\include{chapters/ipa_background}
\include{chapters/ipa_meets_unicode}
\include{chapters/orthography_profiles}

\chapter{Implementation}
\label{implementation}

\section{Available implementations}

To illustrate the practical application of orthography profiles as defined in
Chapter \ref{orthography-profiles}, we have implemented two different versions
of the specifications presented there: one in Python and one in R. These two
implementations have rather different implementation histories, and we are not
100\% sure that they will in all situations give the same results. Also the
perfomance with larger datasets differs, and the code is not always as clean as
we would like it to be. In sum, the two implementations should be considered as
`proof of concept' and not as the final word on the practical application of the
specifications above. In our own experience, the current implementations are
sufficiently fast and stable to be useful for academic practice (e.g. checking
data consistency, or analyzing and transliterating small to medium sized data
sets), but they should probably better not be used for full-scale industry
applications.

This chapter will introduce the implementations, and give practical step-by-step 
guidelines for installing them and using them. Various simple, and sometimes 
somewhat abstract, examples will be discussed to show the different options 
available, and to illustrate the intended usage of orthography profiles. In the 
next chapter we will then discuss a few full-scale examples to illustrate the 
application and the usefulness of orthography profiles.

@@

\noindent Basically, the idea is to use \texttt{write.profile} to produce a
basic orthography profile from some data and then \texttt{tokenize} to apply the
(possibly edited) profile on some data, as exemplified in the next section. This
can of course be performed though R, but additionally there are two more
interfaces to the R code supplied in the \texttt{qlcData} package: (i) bash
executables and (ii) `shiny' webapps.

@@

\noindent For example, a good way to use the executables in a terminal is to
make softlinks (using \texttt{ln}) from the executables to a directory in your
PATH, e.g. to \texttt{/usr/local/bin/}. The two executables are named
\texttt{tokenize} and \texttt{writeprofile}, and the links can be made directly 
by using Rscript to get the paths to the executables within the terminal.

\begin{knitrout}\scriptsize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
# get the paths to the R executables in bash
pathT=`Rscript -e 'cat(file.path(find.package("qlcData"), "exec", "tokenize"))'`
pathW=`Rscript -e 'cat(file.path(find.package("qlcData"), "exec", "writeprofile"))'`

# make softlinks to the R executables in /usr/local/bin
ln -is $pathT $pathW /usr/local/bin
\end{alltt}
\end{kframe}
\end{knitrout}

\noindent After this softlink it should be possible to access the \texttt{tokenize}
function from the shell. Try \texttt{tokenize --help} to test the functionality.

\begin{knitrout}\scriptsize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
tokenize --help
\end{alltt}

@@

\section{Error reporting}
\label{error-reporting}

@@

\noindent The working of error-message reporting can also nicely be illustrated
with this example. Suppose we made an orthography profile with just the two
Latin letters <A> and <B> as possible graphemes, then this profile would not be
sufficient to tokenize the strings. There are graphemes in the data that are not
in the profile, so the tokenization produces an error, which can be used to fix
the encoding. In the example below, we can see that the Cyrillic encoding is
found in the second string of the \texttt{test} input.

@@

\section{Different ways to write a profile}
\label{write-profile}

The function \texttt{write.profile} can be used to prepare a skeleton for an
orthography profile from some data. The preparation of an orthography profile
from some data might sound like a trivial problem, but actually there are
various different ways in which strings can be separated into graphemes by
\texttt{write.profile}. Consider the following example string. The default
settings of \texttt{write.profile} separates the string into Unicode graphemes
according to grapheme clusters (`user-perceived characters', see
Section~\ref{the-unicode-approach}).

@@

% latex table generated in R 3.3.0 by xtable 1.8-2 package
% Thu Jun 30 18:15:48 2016
\begin{table}[H]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Gr. & Freq. & Codepoint & Unicode Name \\ 
  \midrule
Ú & 1 & U+00DA & LATIN CAPITAL LETTER U WITH ACUTE \\ 
  Ú & 1 & U+0055, U+0301 & LATIN CAPITAL LETTER U, COMBINING ACUTE ACCENT \\ 
  Ù & 1 & U+00D9 & LATIN CAPITAL LETTER U WITH GRAVE \\ 
  Ù & 1 & U+0055, U+0300 & LATIN CAPITAL LETTER U, COMBINING GRAVE ACCENT \\ 
  Û & 1 & U+00DB & LATIN CAPITAL LETTER U WITH CIRCUMFLEX \\ 
  Û & 1 & U+0055, U+0302 & LATIN CAPITAL LETTER U, COMBINING CIRCUMFLEX ACCENT \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Profile 1 (default settings, splitting grapheme clusters)} 
\label{tab:profile1}
\end{table}

\noindent Some of these graphemes are single codepoints, others are combinations
of two codepoints. By specifying the splitting separator as empty, it is
possible to split the string into Unicode codepoints.

@@

% latex table generated in R 3.3.0 by xtable 1.8-2 package
% Thu Jun 30 18:15:48 2016
\begin{table}[H]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Grapheme & Frequency & Codepoint & Unicode Name \\ 
  \midrule
́ & 1 & U+0301 & COMBINING ACUTE ACCENT \\ 
  ̀ & 1 & U+0300 & COMBINING GRAVE ACCENT \\ 
  ̂ & 1 & U+0302 & COMBINING CIRCUMFLEX ACCENT \\ 
  U & 3 & U+0055 & LATIN CAPITAL LETTER U \\ 
  Ú & 1 & U+00DA & LATIN CAPITAL LETTER U WITH ACUTE \\ 
  Ù & 1 & U+00D9 & LATIN CAPITAL LETTER U WITH GRAVE \\ 
  Û & 1 & U+00DB & LATIN CAPITAL LETTER U WITH CIRCUMFLEX \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Profile 2 (splitting by codepoints)} 
\label{tab:profile2}
\end{table}

@@

% latex table generated in R 3.3.0 by xtable 1.8-2 package
% Thu Jun 30 18:15:48 2016
\begin{table}[H]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Grapheme & Frequency & Codepoint & Unicode Name \\ 
  \midrule
Ú & 2 & U+00DA & LATIN CAPITAL LETTER U WITH ACUTE \\ 
  Ù & 2 & U+00D9 & LATIN CAPITAL LETTER U WITH GRAVE \\ 
  Û & 2 & U+00DB & LATIN CAPITAL LETTER U WITH CIRCUMFLEX \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Profile 3 (splitting by NFC codepoints)} 
\label{tab:profile3}
\end{table}

% latex table generated in R 3.3.0 by xtable 1.8-2 package
% Thu Jun 30 18:15:48 2016
\begin{table}[H]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Grapheme & Frequency & Codepoint & Unicode Name \\ 
  \midrule
́ & 2 & U+0301 & COMBINING ACUTE ACCENT \\ 
  ̀ & 2 & U+0300 & COMBINING GRAVE ACCENT \\ 
  ̂ & 2 & U+0302 & COMBINING CIRCUMFLEX ACCENT \\ 
  U & 6 & U+0055 & LATIN CAPITAL LETTER U \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Profile 4 (splitting by NFD codepoints)} 
\label{tab:profile4}
\end{table}

@@

\begin{knitrout}\footnotesize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# note that NFC and NFD normalization are identical}
\hlcom{# for unicode grapheme definitions}
\hlstd{profile_5} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(example,} \hlkwc{normalize} \hlstd{=} \hlstr{"NFD"}\hlstd{)}
\hlstd{profile_6} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(example,} \hlkwc{normalize} \hlstd{=} \hlstr{"NFC"}\hlstd{)}
\end{alltt}
\end{kframe}
\end{knitrout}

% latex table generated in R 3.3.0 by xtable 1.8-2 package
% Thu Jun 30 18:15:48 2016
\begin{table}[H]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Gr. & Freq. & Codepoint & Unicode Name \\ 
  \midrule
Ú & 2 & U+0055, U+0301 & LATIN CAPITAL LETTER U, COMBINING ACUTE ACCENT \\ 
  Ù & 2 & U+0055, U+0300 & LATIN CAPITAL LETTER U, COMBINING GRAVE ACCENT \\ 
  Û & 2 & U+0055, U+0302 & LATIN CAPITAL LETTER U, COMBINING CIRCUMFLEX ACCENT \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Profile 5 (splitting by graphemes after NFD)} 
\label{tab:profile5}
\end{table}

% latex table generated in R 3.3.0 by xtable 1.8-2 package
% Thu Jun 30 18:15:49 2016
\begin{table}[H]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Gr. & Freq. & Codepoint & Unicode Name \\ 
  \midrule
Ú & 2 & U+00DA & LATIN CAPITAL LETTER U WITH ACUTE \\ 
  Ù & 2 & U+00D9 & LATIN CAPITAL LETTER U WITH GRAVE \\ 
  Û & 2 & U+00DB & LATIN CAPITAL LETTER U WITH CIRCUMFLEX \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Profile 6 (splitting by graphemes after NFC)} 
\label{tab:profile6}
\end{table}

@@

\begin{knitrout}\scriptsize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
## show help
writeprofile --help
\end{alltt}

@@

\section{Using an orthography profile skeleton}
\label{profile-skeleton}

A common workflow to use these functions is to first make a skeleton for an
orthography profile and then edit this profile by hand. For example, Table
\ref{tab:profile_skeleton1} shows the profile skeleton after a few graphemes have
been added to the file. Note that in this example, the profile is written to the
desktop, and this file has to be edited manually. We simply add a few
multigraphs to the column \texttt{Grapheme} and leave the other columns empty.
these new graphemes are then included in the graphemic parsing.

\begin{knitrout}\footnotesize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# a few words to be graphemically parsed}
\hlstd{example} \hlkwb{<-} \hlkwd{c}\hlstd{(}\hlstr{"mishmash"}\hlstd{,} \hlstr{"mishap"}\hlstd{,} \hlstr{"mischief"}\hlstd{,} \hlstr{"scheme"}\hlstd{)}
\hlcom{# write a profile skeleton to a file}
\hlkwd{write.profile}\hlstd{(example,} \hlkwc{file} \hlstd{=} \hlstr{"~/Desktop/profile_skeleton.txt"}\hlstd{)}
\hlcom{# edit the profile, and then use the edited profile to tokenize}
\hlkwd{tokenize}\hlstd{(example,} \hlkwc{profile} \hlstd{=} \hlstr{"~/Desktop/profile_skeleton.txt"}\hlstd{)}\hlopt{$}\hlstd{strings}
\end{alltt}
\end{kframe}
\end{knitrout}

\begin{knitrout}\footnotesize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
##   originals    tokenized
## 1   shampoo  sh a m p oo
## 2    mishap   m i sh a p
## 3  mischief m i sch ie f
## 4    scheme    sch e m e
\end{verbatim}
\end{kframe}
\end{knitrout}

% latex table generated in R 3.3.0 by xtable 1.8-2 package
% Thu Jun 30 18:15:49 2016
\begin{table}[htb]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Grapheme & Frequency & Codepoint & UnicodeName \\ 
  \midrule
sh &  &  &  \\ 
  ch &  &  &  \\ 
  sch &  &  &  \\ 
  ie &  &  &  \\ 
  oo &  &  &  \\ 
  a & 2 & U+0061 & LATIN SMALL LETTER A \\ 
  c & 2 & U+0063 & LATIN SMALL LETTER C \\ 
  e & 3 & U+0065 & LATIN SMALL LETTER E \\ 
  f & 1 & U+0066 & LATIN SMALL LETTER F \\ 
  h & 4 & U+0068 & LATIN SMALL LETTER H \\ 
  i & 3 & U+0069 & LATIN SMALL LETTER I \\ 
  m & 4 & U+006D & LATIN SMALL LETTER M \\ 
  o & 2 & U+006F & LATIN SMALL LETTER O \\ 
  p & 2 & U+0070 & LATIN SMALL LETTER P \\ 
  s & 4 & U+0073 & LATIN SMALL LETTER S \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Manually edited profile skeleton} 
\label{tab:profile_skeleton1}
\end{table}

For quick solutions, it is also possible to leave out the Unicode information in
the profile skeleton by using the option \texttt{info = FALSE}. It is also
possible not to use a separate file at all, but process everything within R. In
simply situations this is often useful (see below), but in general we prefer to
handle everything through a separately saved orthography profile. This profile
often contains highly useful information that is nicely coded and saved inside
this one file, and can thus be easily distributed and shared. Doing the same as
above completely within R might look as follows:

\begin{knitrout}\footnotesize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# make a profile, just select the column 'Grapheme'}
\hlstd{profile} \hlkwb{<-} \hlkwd{write.profile}\hlstd{(example)[,} \hlstr{"Grapheme"}\hlstd{]}
\hlcom{# extend the profile with multigraphs}
\hlstd{profile} \hlkwb{<-} \hlkwd{c}\hlstd{(}\hlstr{"sh"}\hlstd{,} \hlstr{"ch"}\hlstd{,} \hlstr{"sch"}\hlstd{,} \hlstr{"ie"}\hlstd{,} \hlstr{"oo"}\hlstd{, profile)}
\hlcom{# use the profile to tokenize}
\hlkwd{tokenize}\hlstd{(example, profile)}\hlopt{$}\hlstd{strings}
\end{alltt}
\begin{verbatim}
##   originals    tokenized
## 1   shampoo  sh a m p oo
## 2    mishap   m i sh a p
## 3  mischief m i sch ie f
## 4    scheme    sch e m e
\end{verbatim}
\end{kframe}
\end{knitrout}

To document a specific case of graphemic parsing, it is highly useful to save
all results of the tokenization to file by using the option \texttt{file.out},
for example as follows:

\begin{knitrout}\footnotesize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# save the results to various files}
\hlkwd{tokenize}\hlstd{( example}
         \hlstd{,} \hlkwc{profile} \hlstd{=} \hlstr{"~/Desktop/profile_skeleton.txt"}
         \hlstd{,} \hlkwc{file.out} \hlstd{=} \hlstr{"~/Desktop/result"}
        \hlstd{)}
\end{alltt}
\end{kframe}
\end{knitrout}

This will lead to the following four files being written:

\begin{itemize}

   \item \textsc{result_strings.tsv}:\\ A tab-separated file with the original
         and the tokenized/transliterated strings

   \item \textsc{result_profile.tsv}:\\ A tab-separated file with the
         graphemes with added frequencies of occurence in the data. The lines
         in the file is re-ordered according to the order that resulted from the
         ordering specifications.

   \item \textsc{result_errors.tsv}:\\ A tab-separated file with all original
         strings that contain unmatched parts. Unmatched parts are indicated
         with the character as specified with the option \texttt{missing}. By
         default the character \textsc{double question mark} < ⁇ > at
         \uni{2047} is used. When there are no errors, then this file is 
         absent.

    \item \textsc{result_missing.tsv}:\\ A tab-separated file with the graphemes
          that are missing from the original orthography profile, as indicated in
          the errors. When there are no errors, then this file is absent.

\end{itemize}

\section{Rule ordering}
\label{rule-ordering}

Not yet everything is correct in this graphemic parsing. The sequence <sh> in
`mishap' should not be a digraph, and likewise the sequence <sch> in `mischief'
should of course be separated into <s> and <ch>. One of the important issues to
get the graphemic parsing right is the order in which graphemes are parsed. For
example, currently the grapheme <sch> is parsed before the grapheme <ch>,
leading to <m\ i\ sch\ ie\ f> instead of the intended <m\ i\ s\ ch\ ie\ f>. The
reason that <sch> is parsed before <ch> is that by default longer graphemes are
parsed before shorter ones. Our experience is that in most cases this is
expected behaviour. You can change the ordering by specifying the option
\texttt{ordering}. Setting this option to \texttt{NULL} results in no
preferential ordering, i.e. the graphemes are parsed in the order of the
profile, from top to bottom. Now `mischief' is parsed correctly, but `scheme' is
wrong. So ordering is not the solution in this case.

\begin{knitrout}\footnotesize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# do not reorder the profile}
\hlcom{# just apply the graphemes from top to bottom}
\hlkwd{tokenize}\hlstd{( example}
         \hlstd{,} \hlkwc{profile} \hlstd{=} \hlstr{"~/Desktop/profile_skeleton.txt"}
         \hlstd{,} \hlkwc{ordering} \hlstd{=} \hlkwa{NULL}
        \hlstd{)}\hlopt{$}\hlstd{strings}
\end{alltt}
\end{kframe}
\end{knitrout}

\begin{knitrout}\footnotesize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
##   originals     tokenized
## 1   shampoo   sh a m p oo
## 2    mishap    m i sh a p
## 3  mischief m i s ch ie f
## 4    scheme    s ch e m e
\end{verbatim}
\end{kframe}
\end{knitrout}

There are various additional options for rule ordering implemented. Please check
the help description in R \texttt{help(tokenize)} for more details on the
possible rule ordering specificiations. In summary, there are four different 
ordering options, that can also be combined:

\begin{itemize}

   \item \textsc{size}:\\
         order the lines in the profile by the size of the
         grapheme, largest first. Size is measured by number of Unicode
         characters after normalization as specified in the option
         \texttt{normalize}. For example, <é> has a size of 1 with
         \texttt{normalize = "NFC"}, but a size of 2 with
         \texttt{normalize = "NFD"}.

   \item \textsc{context}:\\
         order the lines by whether they have any context
         specified (see next section). Lines with context coming first. Note
         that this only works when the option \texttt{regex = TRUE} is also chosen.

  \item \textsc{reverse}:\\
        order the lines from bottom to top.

  \item \textsc{frequency}:\\
         order the lines by the frequency with which they
         match in the specified strings before tokenization, least frequent
         coming first. This frequency of course depends crucially on the
         available strings, so it will lead to different orderings when applied
         to different data. Also note that this frequency is (necessarily)
         measured before graphemes are identified, so these ordering frequencies
         are not the same as the final frequencies shown in the output.
         Frequency of course also strongly differs on whether context is used
         for the matching through \texttt{regex = TRUE}.

\end{itemize}

By specifying more than one ordering, these orderings are used to break ties,
e.g. the default setting \texttt{ordering = c("size", "context", "reverse")}
will first order by size, and for those with the same size, it will order by
whether there is any context specified or not. For lines that are still tied
(i.e. the have the same size and both/neither have context) the order will be
reversed compared to the order as attested in the profile. Reversing order can
be useful because hand-written profiles tend to put general rules before
specific rules, which mostly should be applied in reverse order.

It is important to realize that different ordering of the rules does not have
so-called `feeding' and `bleeding' effects as known from finite-state rewrite
rules.\footnote{`Bleeding' is the effect that the application of a rule changes
the string, so as to prevent a next rule to apply. `Feeding' is the opposite: a
specific rule will only be applied becuase a previous rule changed the string
already. The interaction of rules with such `feeding' and `bleeding' effects is
extremely difficult to predict.} The graphemic parsing advocated here works
crucially different from rewrite rules in that there is nothing being rewritten:
each line in an orthography profile specifies a grapheme to be `captured' in the 
string. All lines in the profile are processed in a specifed order (as determined
by the option \texttt{ordering}). At the processing of a specific line, all 
matching graphemes in the data are marked as `captured', but not changed. 
Captured parts cannot be captured again, but they can still be used to match 
contexts of other lines in the profile. Only when all lines are processed the 
captured graphemes are separated (and possibly transliterated). In this way the 
result of the application of the `rules' is rather easy to predict.

To document the process of graphemic tokenization, it is strongly encouraged to

\section{Contextually specified graphemes}
\label{contextual-specification}

To refine a profile, it is also possible to add graphemes with contextual
specifications. An orthography profile can have columns called \texttt{Left} and
\texttt{Right} to specify the context in which the Grapheme is to be
separated.\footnote{The columns names \textttf{Left}, \textttf{Right} and
\textttf{Grapheme} are currently hard coded, so exactly these column names
should be used for these effects to take place. The position of the columns in
the profile is unimportant. So the column \textttf{Left} can occur anywhere.} We
are adding an extra line to the profile specifying that <s> is a grapheme when
it occurs after <mi>. These contextually specified graphemes are based on
regular expressions (so you can actually use regular expressions in the
description of the context). For these to be included in the graphemic parsing 
we have to specify the option \texttt{regex = TRUE}. This contextually specified 
grapheme should actually be handled first, so we could try \texttt{ordering = 
NULL}. However, we can also explicitly specifiy that rules with contextual 
information should be applied first by using \texttt{ordering = "context"}. That 
gives the right results for this toy example.

\begin{knitrout}\footnotesize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# add a contextual grapheme, and then use the edited profile to tokenize}
\hlkwd{tokenize}\hlstd{( example}
         \hlstd{,} \hlkwc{profile} \hlstd{=} \hlstr{"~/Desktop/profile_skeleton.txt"}
         \hlstd{,} \hlkwc{regex} \hlstd{=} \hlnum{TRUE}
         \hlstd{,} \hlkwc{ordering} \hlstd{=} \hlstr{"context"}
        \hlstd{)}\hlopt{$}\hlstd{strings}
\end{alltt}
\end{kframe}
\end{knitrout}

\begin{knitrout}\footnotesize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
##   originals     tokenized
## 1   shampoo   sh a m p oo
## 2    mishap   m i s h a p
## 3  mischief m i s ch ie f
## 4    scheme    s ch e m e
\end{verbatim}
\end{kframe}
\end{knitrout}
